Endogenous Macrodynamics in Algorithmic Recourse

### 2022-11-03 dev/submissions/satml/revision.qmd
cut example

>>>>

Thanks - we have now added a subsection to the section on limitations:

"In this work we have focused on popular counterfactual generators that do not incorporate any causal knowledge. The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome. The implementation of such changes is typically described as gaming [36]. Preventing such gaming may help to avoid the macrodynamics we have pointed to in this work. Future work would therefore likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al. [13]."

Does this sound reasonable? By adding it as a limitation we make it clear that this is in fact a shortcoming of our current work, which should help with toning down the narrative. Beyond that, were there any specific points in the paper that struck you as too boastful?

We also thought about potentially linking this story on causality back to one of the examples, but have so far struggled to find any obvious non-causal attribute and are anyway struggling with space. We thought of this: "In the context of Example I.2, the generators may, for example, propose that students improve their results on some aptitude test, that is recognized as a signal by the university’s hiring committee. Changes to those test scores can be expected to affect the underlying machine learning model even if there is no causal link between test outcomes and students’ chances of graduating." Another example we could think of is motivational statements that may or may not be causally linked to student performance.

As for your point on retraining, ideally, what we'd want is to actually intervene and retrain on observed outcomes after the intervention. We would then expect that the model would "catch the fact that at least some people who relied on AR actually failed to experience the predicted outcome (e.g., repayment or completing their degree)". This would set the baseline for a very interesting study of the actual causal effect of algorithmic recourse on actual (as opposed to predicted) outcomes. But since here we're dealing with observational real-world data, we were more limited in our analysis.

To be clear, during retraining we consider new predicted labels as new observed labels. The alternative would be to retrain with existing labels in which case we would indeed also expect the decision boundary to shift in the direction of the target class. Both approaches are simplifications and not quite optimal (which we recognize in the limitations section). Our approach is simplified, because we assume that the agent in charge of the model accepts the new labels as ground truth. The alternative approach is also not optimal, because it assumes that changes to features have no affect on actual outcomes.

Again, since we never actually observe the outcome after implementation of recourse, we are limited here. But having thought about this again now, an interesting idea for future work might be to generate synthetic data from a simple causal model.

Thanks again!

### 2022-11-03 paper/paper.tex
cut example

In this work we have focused on popular counterfactual generators that do not incorporate any causal knowledge. The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome. In the context of Example \ref{exm:student}, the generators may, for example, propose that students improve their results on some aptitude test, that is recognized as a signal by the university's hiring committee. Changes to those test scores can be expected to affect the underlying machine learning model even if there is no causal link between test outcomes and students' chances of graduating. The implementation of such changes is typically described as \textbf{gaming} \protect\hyperlink{ref-miller2020strategic}{{[}36{]}}. Preventing such gaming may help to avoid the macrodynamics we have pointed to in this work. Future work would therefore likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al. \protect\hyperlink{ref-karimi2021algorithmic}{{[}13{]}}.

>>>>

In this work we have focused on popular counterfactual generators that do not incorporate any causal knowledge. The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome. The implementation of such changes is typically described as \textbf{gaming} \protect\hyperlink{ref-miller2020strategic}{{[}36{]}}. Preventing such gaming may help to avoid the macrodynamics we have pointed to in this work. Future work would therefore likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al. \protect\hyperlink{ref-karimi2021algorithmic}{{[}13{]}}.

### 2022-11-03 paper/sections/limitations.rmd
cut example

In this work we have focused on popular counterfactual generators that do not incorporate any causal knowledge. The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome. In the context of Example \@ref(exm:student), the generators may, for example, propose that students improve their results on some aptitude test, that is recognized as a signal by the university's hiring committee. Changes to those test scores can be expected to affect the underlying machine learning model even if there is no causal link between test outcomes and students' chances of graduating. The implementation of such changes is typically described as **gaming** @miller2020strategic. Preventing such gaming may help to avoid the macrodynamics we have pointed to in this work. Future work would therefore likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al. @karimi2021algorithmic.

>>>>

In this work we have focused on popular counterfactual generators that do not incorporate any causal knowledge. The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome. The implementation of such changes is typically described as **gaming** @miller2020strategic. Preventing such gaming may help to avoid the macrodynamics we have pointed to in this work. Future work would therefore likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al. @karimi2021algorithmic.
